---
title: "Reproducibility in R Bootcamp 2025"
author: "Eric Salomaki"
date: "2025-05-22"
output: html_document
---

```{r}
library(here)
library(kableExtra)
library(patchwork)
library(pheatmap)
library(ggupset)
library(DESeq2)
library(pathview)
library(apeglm)
library(IHW)
library(ggplot2)
library(ggrepel)
library(qs2)
library(sva)
```

## here sets a location relative to the R script
```{r}
here::i_am("reproducibility_in_r.Rmd")
```

## lets set here to the top directory of the repo
```{r}
here::i_am("notebook/reproducibility_in_r.Rmd")
```


## now we can easily read in some data
```{r}
counts_matrix <- read.csv(here("data","sample_data.tsv"), sep = "\t", header = TRUE, row.names = 1)
counts_matrix
```

## And read in the sample metadata
```{r}
metadata <- read.csv(here("data","sample_metadata.tsv"), sep = "\t", header = TRUE, row.names = 1)
metadata
```

## Now make DESeq dataset
```{r}
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = metadata,
                              design = ~ condition)
dds
```

## And run DESeq2
```{r}
dds <- DESeq(dds)
res <- results(dds)
rld <- rlog(dds, blind = FALSE)
res_shrunk <- lfcShrink(dds, coef = "condition_Treatment_vs_Control")
```

## Save your DESeq outputs (or any analysis files that take a long time to run)
```{r}
# Save as a qs2 object using qs2 package fucntions
qs_save(dds, here("data","DESeq_dds.qs2"))
# Read a qs2 object and validate file integrity
dds <- qs_read(here("data","DESeq_dds.qs2"), validate_checksum = TRUE) 

# For small files, base R functions work similarly
# Save as a RDS object using base R fucntions
saveRDS(dds, here("data","DESeq_dds.RDS"))
# Read a a RDS object, lacks the ability to validate file integrity
dds2 <- readRDS(here("data","DESeq_dds.RDS")) 
```


## Plot PCA of data
```{r}
plotPCA(rld)
```


## Make a sample distance plot
```{r}
sampleDists <- dist(t(assay(rld)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- colnames(rld)
colnames(sampleDistMatrix) <- colnames(rld)

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         main = "Sample-to-sample distances")
```

### Make an MA plot
```{r}
plotMA(res_shrunk, main = "MA Plot (shrunken LFC)", ylim = c(-3, 3))
```



### Make a volcano plot
```{r}
res_df <- as.data.frame(res_shrunk)
res_df$gene <- rownames(res_df)
res_df$significant <- with(res_df, padj < 0.05 & abs(log2FoldChange) > 2)

ggplot(res_df, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = significant), alpha = 0.7) +
  scale_color_manual(values = c("black", "red")) +
  geom_vline(xintercept = c(-2, 2), linetype = "dashed", color = "gray") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray") +
  geom_text_repel(data = subset(res_df, significant),
                  aes(label = gene),
                  size = 3,
                  max.overlaps = 20,
                  box.padding = 0.3,
                  point.padding = 0.2) +
  theme_minimal() +
  labs(title = "Volcano Plot", x = "Log2 Fold Change", y = "-Log10 p-value")
```


## Finally we will print the session info
```{r}
sessionInfo()
```




